[Re] Predicting metabolic biomarkers of human inborn errors of metabolism

### 2018-02-20 article/Mondeel_Ogundipe_Westerhoff-2017.tex
RE: Aaron 2.0
- Small text changes
- medium is not -10 inward not -1. This results in correct predictions for Figure 3.
- Include text and code to show result for -1 related to Figure 3
- Fix ordering Figure 1

Shlomi et al. illustrated the method first on a virtual metabolic map
and further validated their technique using the reconstruction of the
human metabolic map Recon1 \autocite{Duarte2007} empirically with
respect to known biomarkers, but did not provide any formal proof. For a
set of 17 IEMs that affect amino acid metabolism, they found a 56\%
fraction of the biomarkers that are correctly predicted and a fraction
of 76\% of predicted biomarkers that were predicted correctly in terms
of whether they increased or decreased in concentration. Thiele et al.
\autocite{Thiele2013} later found an even higher precision using
essentially the same FVA methodology on an updated version of the
metabolic reconstruction.

@@

Exchange reactions are reactions in the metabolic network that allow
import and/or export of metabolites from the metabolic map. Exchange
reactions are represented by non-mass-balanced pseudoreactions, e.g.
\(X \leftrightarrow \varnothing\), where \(X\) is understood to
originate from or efflux to the biofluids. Positive flux indicates net
secretion of \(X\) by the cell and negative flux indicates the net
uptake of \(X\).

@@

\begin{align}
\text{Maximize } &\pm v_i, \text{ such that for all } k \nonumber \\
\bm{S} \bm{v} &= \bm{0} \nonumber \\
Z  &\geq \phi Z_{\text{max}} \nonumber \\
\alpha_k \leq & ~ v_k \leq \beta_k 
\end{align}

Here \(\bm{v}\) is the column vector of fluxes representing all
reactions in the model, \(v_i\) is the flux of the exchange reaction of
a given boundary metabolite \(i\), \(\alpha_k\) is the, possibly
negative, lower bound for reaction \(k\) and similarly \(\beta_k\) is
the, possibly negative, upper bound for reaction \(k\). The reaction
bounds are the \(V_{max}\)'s of the reactions and are part of the
metabolic network definition, typically they are set to either \(0\) or
\(\pm 1000\) if the true, biological, \(V_{max}\)'s are not known. These
values allow specification of irreversible reactions by setting the
lower bound (or upper bound) to zero. The index \(i\) specifies a single
reaction that is to be maximized, whereas the index \(k\) is used to
index the bound constraints that exist for each reaction, including
\(v_i\). \(v_i\) may be maximized in the forward (positive flux) or the
reverse (negative flux) direction if allowed by the bounds on \(v_i\).
This is indicated by the \(\pm\) symbol. \(\bm{S}\) is the stoichiometry
matrix for the metabolic network of size \(m \times r\) where \(m\) is
the number of metabolites and \(r\) the number of reactions. The
numeric, typically integer, elements of \(\bm{S}\) represent the
stoichiometry coefficients of each metabolite \(m\) in each reaction
\(r\). Consequently, for large networks, this matrix is sparse (has many
zero entries). \(Z = \bm{c}^T \bm{v}\) is the objective function defined
for the map, entailing a linear combination, defined by column vector
\(\bm{c}\), of one or more reactions. The vector \(\bm{c}\) is an
indicator of objective reactions, i.e.~it contains a value of \(1\) in
rows corresponding to fluxes that are to be included in the objective,
and zeros in all other rows. Due to the vector multiplication
\(\bm{c}^T \bm{v}\), the objective \(Z\) sums the fluxes of the
reactions that correspond to rows containing a value of \(1\) in
\(\bm{c}\). \(Z_{\text{max}}\) denotes the maximal value of the
objective function. We define \(0 \leq \phi \leq 1\) to be an arbitrary
minimal fraction of the maximal objective function value that has to be
achieved. When \(\phi=1\) we ask for the range of flux allowed for
reaction \(v_i\) while maintaining the maximal value of the linear
combination of objective fluxes. When setting \(\phi = 0\), one is
asking for the allowable flux range through reaction \(v_i\) regardless
of the value of the objective function (although the sign has to be
maintained). The latter, i.e. \(\phi=0\), is the case considered in the
method by Shlomi et al. The choice of \(\phi\) matters because when
\(\phi \neq 0\) we are requiring flux to flow through the set of
objective reactions (unless \(Z_{\text{max}} = 0\)). This may entail a
forced directionality through reactions that are required to ultimately
produce flux in the objective reactions. These additional limitations in
the freedom of the flux pattern may subsequently impact the biomarker
predictions. The approach discussed here entails predictions purely
based on the network topology and therefore sets \(\phi = 0\).

@@

Returning to the example of phenylketonuria, we simulated this for Recon
2.2 using the synchronous approach. Column 3 in Table
{[}tbl:PKU\_results{]} shows that the result now agrees with the Recon 1
asynchronous predictions. Using Recon 1, we came to the correct
prediction because it contains only one reaction (`PHETHPTOX2') linked
to the PKU gene, i.e.~only phenylalanine hydroxylase (Entrez ID: 5053,
HGNC:8582):

\begin{align*}
&\text{(6R)-L-erythro-5,6,7,8-tetrahydrobiopterin} + \text{L-phenylalanine} + \text{O}_2 \\
&\leftrightarrow \\
&\text{4a-hydroxy-L-erythro-5,6,7,8-tetrahydrobiopterin} + \text{L-tyrosine}
\end{align*}

@@

\begin{align*}
&\text{(6R)-L-erythro-5,6,7,8-tetrahydrobiopterin} + \text{L-phenylalanine} + \text{O}_2 \\
&\leftrightarrow \\
&\text{6,7-dihydrobiopterin} + \text{L-tyrosine} + \text{H}_2\text{O}.
\end{align*}

@@

The original publication's Figure 1, panel A and B, exemplified the
biomarker prediction method for an illustrative network. It is
reproduced here as Figure ~\ref{fig:Shlomi2009_figure_1}. The simple
nature of the example serves to explain the reasoning behind the method
and also helped us validate our implementation of the method. This
repository provides an SBML file of the network, the Jupyter notebook
that generates it and a Jupyter notebook reproducing the biomarker
prediction.

@@

It is worthwhile to note, with an eye on the sensitivity of this
approach to the various settings, that we needed to set the medium
composition very specifically. Shlomi et al. did not report on their
chosen medium in their publication but we were able to reproduce their
results by setting all inward flux bounds to \(-1\) and all outward flux
bounds to \(1000\).

@@

\begin{figure}
\centering
\includegraphics{Figures/Reproduction_Shlomi_fig_3}
\caption{Reproduction of Figure 3 in \autocite{Shlomi2009}. (A)
Illustration of the subnetwork of Recon 1 relevant to the effect of
homocystinuria on the metabolism and transport of methionine. Nodes
represent metabolites and edges represent reactions. For simplicity,
only abbreviations of metabolite names are given in the figure. The full
names are: L-Methionine (met), S-Adenosyl-L-methionine (amet),
S-Adenosyl-L-homocysteine (ahcys), L-Homocysteine (hcys), L-Serine
(ser), L-Cystathionine (cyst), L-Cysteine (cys),
S-Adenosylmethioninamine (ametam), 5-Methylthioadenosine (5mta),
5-Methylthio-5-deoxy-D-ribose 1-phosphate (5mdr1p),
5-Methylthio-5-deoxy-D-ribulose 1-phosphate (dmdru1p),
2,3-diketo-5-methylthio-1-phosphopentane (dkmpp),
2-keto-4-methylthiobutyrate (2kmb). Metabolites with a green background
color participate in additional reactions not shown here. Reactions are
indicated by their identifiers in Recon 1 and with a corresponding
enzyme commision number if available. We highlighted the mutations for
Homocystinuria (dysfunctional CBS) and hypermethioninemia (dysfunctional
AHCY). (B) Prediction of concentration changes of methionine in
homocystinuria and S-adenosylhomocysteine hydrolase deficiency based on
the biomarker prediction algorithm.}\label{fig:Shlomi2009_figure_3}
\end{figure}

Figure 3 in the original publication entailed an in-depth look at the
results highlighted in Figure 2 for methionine metabolism. We reproduced
this Figure as Figure ~\ref{fig:Shlomi2009_figure_3}. The numerical
results illustrated in ~\ref{fig:Shlomi2009_figure_3}B may be reproduced
using the notebook `Reproduce\_figure2\_Shlomi2009', the detailed
results may be inspected using the last cell of that notebook. We note
that in the original publication the WT interval for the CBS IEM was
drawn as an interval whereas our results indicate an interval of
\([-1,-1]\).

@@

>>>>

Shlomi et al. illustrated the method first on a virtual metabolic map
and further validated their technique using the reconstruction of the
human metabolic map Recon1 \autocite{Duarte2007} empirically with
respect to known biomarkers, but did not provide any formal proof. For a
set of 17 IEMs that affect amino acid metabolism, they found 56\% of the
biomarkers that are correctly predicted and 76\% of predicted biomarkers
that were predicted correctly in terms of whether they increased or
decreased in concentration. Thiele et al. \autocite{Thiele2013} later
found an even higher precision using essentially the same FVA
methodology on an updated version of the metabolic reconstruction.

@@

Exchange reactions are reactions in the metabolic network that allow
import and/or export of metabolites from the metabolic map. Exchange
reactions are represented by non-mass-balanced pseudoreactions, e.g.
\(X \leftrightarrow \varnothing\). Positive flux indicates net secretion
of \(X\) by the cell and negative flux indicates the net uptake of
\(X\).

@@

\begin{align}
\text{Min } v_i \text{ or Max } &v_i, \text{ such that for all } k \nonumber \\
\bm{S} \bm{v} &= \bm{0} \nonumber \\
Z  &\geq \phi Z_{\text{max}} \nonumber \\
\alpha_k \leq & ~ v_k \leq \beta_k 
\end{align}

Here \(\bm{v}\) is the column vector of fluxes representing all
reactions in the model, \(v_i\) is the flux of the exchange reaction of
a given boundary metabolite \(i\), \(\alpha_k\) is the, possibly
negative, lower bound for reaction \(k\) and similarly \(\beta_k\) is
the, possibly negative, upper bound for reaction \(k\). The reaction
bounds are the \(V_{max}\)'s of the reactions and are part of the
metabolic network definition, typically they are set to either \(0\) or
\(\pm 1000\) if the true, biological, \(V_{max}\)'s are not known. These
values allow specification of irreversible reactions by setting the
lower bound (or upper bound) to zero. The index \(i\) specifies a single
reaction that is to be maximized, whereas the index \(k\) is used to
index the bound constraints that exist for each reaction, including
\(v_i\). \(v_i\) may be maximized in the forward (positive flux) or the
reverse (negative flux) direction if allowed by the bounds on \(v_i\).
For a network with \(m\) metabolites and \(r\) reactions, i.e.~fluxes,
\(\bm{S}\) is the stoichiometry matrix for the network of size
\(m \times r\). The numeric, typically integer, elements of \(\bm{S}\)
represent the stoichiometry coefficients of each metabolite \(m\) in
each reaction \(r\). \(Z = \bm{c}^T \bm{v}\) is the objective function
defined for the map, entailing a linear combination, defined by column
vector \(\bm{c}\), of one or more reactions. The vector \(\bm{c}\) is an
indicator of objective reactions, i.e.~it contains a value of \(1\) in
rows corresponding to fluxes that are to be included in the objective,
and zeros in all other rows. Due to the vector multiplication
\(\bm{c}^T \bm{v}\), the objective \(Z\) sums the fluxes of the
reactions that correspond to rows containing a value of \(1\) in
\(\bm{c}\). \(Z_{\text{max}}\) denotes the maximal value of the
objective function. We define \(0 \leq \phi \leq 1\) to be an arbitrary
minimal fraction of the maximal objective function value that has to be
achieved. When \(\phi=1\) we ask for the range of flux allowed for
reaction \(v_i\) while maintaining the maximal value of the linear
combination of objective fluxes. When setting \(\phi = 0\), one is
asking for the allowable flux range through reaction \(v_i\) regardless
of the value of the objective function (although the sign has to be
maintained). The latter, i.e. \(\phi=0\), is the case considered in the
method by Shlomi et al. The choice of \(\phi\) matters because when
\(\phi \neq 0\) we are requiring flux to flow through the set of
objective reactions (unless \(Z_{\text{max}} = 0\)). This may entail a
forced directionality through reactions that are required to ultimately
produce flux in the objective reactions. These additional limitations in
the freedom of the flux pattern may subsequently impact the biomarker
predictions. The approach discussed here entails predictions purely
based on the network topology and therefore sets \(\phi = 0\).

@@

Returning to the example of phenylketonuria, we simulated this for Recon
2.2 using the synchronous approach. Column 3 in Table
~\ref{tbl:PKU_results} shows that the result now agrees with the Recon 1
asynchronous predictions. Using Recon 1, we came to the correct
prediction because it contains only one reaction (`PHETHPTOX2') linked
to the PKU gene, i.e.~only phenylalanine hydroxylase (Entrez ID: 5053,
HGNC:8582):

\begin{gather*}
\text{(6R)-L-erythro-5,6,7,8-tetrahydrobiopterin} + \text{L-phenylalanine} + \text{O}_2 \\
\updownarrow \\
\text{4a-hydroxy-L-erythro-5,6,7,8-tetrahydrobiopterin} + \text{L-tyrosine}
\end{gather*}

@@

\begin{gather*}
\text{(6R)-L-erythro-5,6,7,8-tetrahydrobiopterin} + \text{L-phenylalanine} + \text{O}_2 \\
 \updownarrow \\
\text{6,7-dihydrobiopterin} + \text{L-tyrosine} + \text{H}_2\text{O}.
\end{gather*}

@@

The original publication's Figure 1, panel A and B, exemplified the
biomarker prediction method for an illustrative network. It is
reproduced here as Figure ~\ref{fig:Shlomi2009_figure_1}. The simple
nature of the example serves to explain the reasoning behind the method
and also helped us validate our implementation of the method. This
repository provides an SBML file of the network, the Jupyter notebook
that generates it and a Jupyter notebook reproducing the biomarker
prediction. The image itself was produced using Adobe Illustrator.

@@

It is worthwhile to note, with an eye on the sensitivity of this
approach to the various settings, that we needed to set the medium
composition very specifically. Shlomi et al. did not report on their
chosen medium in their publication but we were able to reproduce their
results by setting all inward flux bounds to \(-10\) and all outward
flux bounds to \(1000\).

@@

\begin{figure}
\centering
\includegraphics{Figures/Reproduction_Shlomi_fig_3}
\caption{Reproduction of Figure 3 in \autocite{Shlomi2009}. (A)
Illustration of the subnetwork of Recon 1 relevant to the effect of
homocystinuria on the metabolism and transport of methionine. Nodes
represent metabolites and edges represent reactions. For simplicity,
only abbreviations of metabolite names are given in the figure. The full
names are: L-Methionine (met), S-Adenosyl-L-methionine (amet),
S-Adenosyl-L-homocysteine (ahcys), L-Homocysteine (hcys), L-Serine
(ser), L-Cystathionine (cyst), L-Cysteine (cys),
S-Adenosylmethioninamine (ametam), 5-Methylthioadenosine (5mta),
5-Methylthio-5-deoxy-D-ribose 1-phosphate (5mdr1p),
5-Methylthio-5-deoxy-D-ribulose 1-phosphate (dmdru1p),
2,3-diketo-5-methylthio-1-phosphopentane (dkmpp),
2-keto-4-methylthiobutyrate (2kmb). Metabolites with a green background
color participate in additional reactions not shown here. Reactions are
indicated by their identifiers in Recon 1 and with a corresponding
enzyme commision number if available. We highlighted the mutations for
Homocystinuria (dysfunctional CBS) and hypermethioninemia (dysfunctional
AHCY). (B) Prediction of concentration changes of methionine in
homocystinuria and S-adenosylhomocysteine hydrolase deficiency based on
the biomarker prediction algorithm. The wild-type (WT) interval is
indicated in black, the mutant interval in red. Each interval is
associated with a numeric interval predicted by the biomarker prediction
algorithm and with a qualitative prediction for the level of the
metabolite in the biofluids (see methods). H.C. is an abbreviation for a
highly confident biomarker, see main
text.}\label{fig:Shlomi2009_figure_3}
\end{figure}

Figure 3 in the original publication entailed an in-depth look at the
effect of inborn errors in AHCY and CBS on methionine metabolism. We
reproduced this Figure as Figure ~\ref{fig:Shlomi2009_figure_3}. We used
Adobe Illustrator to redraw the network from the original publication
and based the intervals in panel B on the numerical results the
algorithm predicts. The qualitative results are already contained in
~\ref{tbl:amino_acid_iems}. The numerical results illustrated in
~\ref{fig:Shlomi2009_figure_3}B may be reproduced using the notebook
\emph{Reproduce\_figure\_2\_and\_3\_Shlomi2009.ipynb}. For clarity we
included a section in the notebook that details the results relevant to
Figure ~\ref{fig:Shlomi2009_figure_3}B.

We note that in early versions of this reproduction we were utilising a
medium that allowed influx of \(-1\) for all medium components, as
opposed to \(-10\) now. It turns out that a setting of \(-1\) does
accurately reproduce the results for Figure 2, but not for Figure 3. For
the CBS IEM the WT interval would correspond to \([-1,-1]\) and would
therefore be a point instead of a wide interval as Shlomi et al. draw
it. We included this simulation in the last cell of the notebook
\emph{Reproduce\_figure\_2\_and\_3\_Shlomi2009.ipynb} This is another
example of the inherent sensitivity of the approach to various settings
in the algorithm.

@@

We hope this reference implementation is useful to the community and can
serve as a starting point for further implementations of this approach
and serves as a warning of the accompanying sensitivities to various
algorithm settings we have lain out in this manuscript.
